---
title: "Script to convert Snapshot-Digikam records to Darwin Core (DwC) Format for FBIP"
author: "Lain E. Pardo"
date: "6/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code converts a report manually generated with DIGIKAM to the Darwin Core (DwC) format widely used for biodiversity records and by FBIP in South Africa. Here we are including ALL columns from original reports as extra information.

We need three files that must be uploaded in order to retrieve the information (see below). These files must be in the same folder for the function to run and the object name should not change. Note that a few FBIP column names might not match correctly the DwC columns codes, so a further processing might be needed when submitting the records to the Catalogue of Life.

files:   
1. sp_rec: the final report we amended to be converted into the DwC format (FBIP), (No        independence between records is considered 
2. metadata: this is the file with all metadata for all sites, we can use the                 "1_Metadata_all_fixed_as_snapshot_TEMP_CSV.csv"
3. traits: we need some taxonomic information from species so we can use the traits file      "+traits_in_sp_records_updated")

*elevation, province and other metadata were retrieved in previous process (see scripts, and readme of metadata for source)

## call libraries
```{r, message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)
```

## load files

```{r, warning= FALSE, message= FALSE}

        
sp_rec <- read_csv("data_in/GOE_sp_report_digikam_2020-08-25_fin.csv") 

metadata <- read_csv("data_in/1_Metadata_all_fixed_as_snapshot_TEMP_CSV.csv")

traits <- read_csv("data_in/+traits_in_sp_records_updated.csv")


```

Most likely your files are in our repositories or other folders in your PC, in this case use you need to set your path, so use the below sample code:

```{r, eval = FALSE }
#knitr::opts_knit$set(root.dir = "C:/Users/admin/Dropbox/Postdoc_NMU/snapshot #safari/reports_Jan")  #set up another working dir
# don´t forget to delete eval if using this chunk
```

```{r, eval = FALSE, warning= FALSE, message= FALSE}
## load files

# path_sp <- "C:/Users/admin/Dropbox/Postdoc_NMU/snapshot safari/data_available_survey_docs/Spp_reports_amended_50%_agreement/1_last_amend_jan/"
# rep <- "MTZ/MTZ_S2_full_report_0-50%_agreement_corrected_JV.csv"
#            
# sp_rec <- read.csv(paste0(path_sp, rep)) #, head = TRUE, sep=";") # modify accordingly
# 
# metadata <- read_csv("data_in/1_Metadata_all_fixed_as_snapshot_TEMP_CSV.csv")
# 
# traits <- read_csv("data_in/+traits_in_sp_records_updated.csv")
# 
# don´t forget to delete eval if using this chunk
```

since we are now working from a "non-clean" report I have to do some previous wrangling before the actual function for DwC format (this is not an issue if working from clean independent reports)

## Data preparation

be mindful that this code eliminates birds and false triggers, if birds are need set FALSE the rm_bycatch argument (but probably some things won´t work e.g. scientific names)

```{r}
source("1_SNAPSHOT_source_functions.R")

#IF ROAMINGS SITE TAKE THE # OUT AND RUN CHUNK WITH THE FOLLOWING LINE FIRST

newrep <- digi_snap2(sp_rec, sitecode = "GOE")  # 1 st arg = object with the original digikam report format, second the code need for new col

newrep <- digi_snap2(sp_rec, sitecode = "MAR")  # 1 st arg = object with the original digikam report format, secont the code need for new col


# before cols_snap_std we need to keep the use site_ID and rename it site
newrep$site <- newrep$site_ID

#reps1 <- cols_snap_std(newrep) 
reps1 <- newrep
# 1_ standardize names of species

df <- standardise_names(reps1)
unique(df$question__species)

# 2_ Clean table (rename species, cast column types)
reps2 <- df %>%
  mutate(DateTimeOriginal = paste(capture_date_local, capture_time_local),
         question__count_max = question__count_median,
         question__count_min = question__count_median)


#----------
df.clean <- clean_table(reps2, rm_bycatch = TRUE, pres_abs = FALSE)


#df.clean_NO_BYCATCH <- clean_table(df, rm_bycatch = TRUE, pres_abs = FALSE)
unique(df.clean$question__species)

sp_rec <- df.clean


```



also there could be some discrepancies between names, some reports would have for example generic names for some specie e.g. zebra instead of zebraburchells, zebramountain as in AUG.


1. if we are sure all species are one or the other do:
```{r}

#e.g.
#unique(sp_rec$question__species)
#class(sp_rec$question_question__species)
#sp_rec$question__species <- as.character(sp_rec$question__species)
#change as 
#sp_rec$question__species[sp_rec$question__species == "zebra"] <- "zebraburchells"
#unique(sp_rec$question__species)
#g <- filter(sp_rec, question__species ==  "genet(rusty)") 
```

<<<<<<< HEAD
2. if we don´t know then leave it but we will need to add a name in the traits doc to retrieve it only that it won´t have scientific name! so do it from file add "sp" in scientifc name
=======
2. if we don´t know then leave it but we will need to add a name in the traits doc to retrieve it only that it won´t have scientific name! so do it from file add "nn" in scientifc name


```{r}
```


## now the function for DwC:

```{r}
# from species records we need:
# 1. add the same value of Photo.ID to new col "Catalogue Number", "Record Number" "Record Number
# 2. add the same values of "Number.of.Individuals" to new col "Individual Count"
# 4. add same value of "Photo.Date" into new col "Date Identified"
# 3. split photo.date into 3 cols: "Year", Month, Day
# 6. add new cols with the same value for all sp records: 

DarwinCore <- function(sp_rec, metadata, traits){
sp_rec1 <- sp_rec %>% 
  mutate("Catalogue Number" = sp_rec$subject_id,
         "Record Number" = sp_rec$subject_id,
         "Individual Count" = sp_rec$question__count_median)


sp_rec1$capture_date_local <- as.Date(sp_rec1$capture_date_local)

sp_rec2<- sp_rec1 %>% 
  mutate("Date Identified" = capture_date_local,
         "Institution ID" = "Nelson Mandela University",
         "Basis of Record" = "Camera-Trap photo",
         "Recorded By" = "Snapshot Safari SA & Wildlife Ecology Lab",
         "Preparations" = "Photo-digital",
         "geodetic Datum" = "WGS84",
         "IdentifiedBy" = "Snapshot Safari SA & Wildlife Ecology Lab",
         "Sampling Protocol" = "Camera-Trapping") %>% 
   separate("Date Identified", c("Year", "Month", "Day"), sep = "-", remove = FALSE)

# # 4_ need standard name for sites including col "site" and "season"
# 
# df1 <- separate(sp_rec2,season, c("code_loc", "season1"), sep = "_", remove =FALSE)  
# 
# 
# #  back to main function
# sp_rec2 <-unite(df1, Camera.Site, code_loc, site, sep = "_",remove =FALSE) #
# 
# unique(sp_rec2$Camera.Site)

# from metadata I need to:--------------------

# 5. take RSA-vegetation to habitat in (into above df), use site id for this I guess
# 6. take cols country, province, municipality and elevation, as they are into each spp above
# 7. take official name and use it in new col named "Locality"
# 8. take the coordinates and put them as a new cols "Decimal Latitude", "Decimal Longitude"
 
#so let´s select variables:
#insert col with reserve code

vars_need <- select(metadata,5,9,10,22, 47:53)

vars_need1 <- vars_need %>% rename(          
  Camera.Site = Camera.Site.Concatenate,
  Locality = Official.Name,
  "Decimal Latitude" = Lat_Y,
  "Decimal Longitude" = Long_X,
  Habitat = Vegetation,
  "Field Notes" = Habitat.Descriptor)
unique(vars_need1$Camera.Site)

#let´s make a left join 
#but first rename site in sp_rec2
sp_rec2 <- rename(sp_rec2,Camera.Site = site)
sp_rec3 <- left_join(sp_rec2, vars_need1, by = "Camera.Site") #working
#names(sp_rec2)
#names(vars_need1)


# now add traits columns --------------------------------------------------

#select those needed

traits1<- select(traits, 4,5,6,7:9)


#use traits to filter sp3 to get only mammals

#add new cols
traits2<- traits1 %>% 
  mutate(Kingdom = "Animalia",
         Phylum = "Chordata",
         Class = "Mammalia Linnaeus, 1758",
         "Vernacular Name" = Common.Name) %>%  # of traits not sp_rec
  rename("Scientific Name Authorship" = Scientific.Name.Authorship) %>% 
  unite("full Scientific Name+author details", Scientific.Name, "Scientific Name Authorship", sep = ", ", remove =FALSE) %>% 
  separate(Scientific.Name, c("Genus", "Species", "Infraspecific Epithet"), sep = "[^[:alnum:]]+", remove = FALSE)
 
#in sp_rec3 change Common.Name for Snapshot.Name to join
#better repeat the column otherwise this name is lost and we want it in the end...
sp_rec3$Snapshot.Name <- sp_rec3$question__species
#sp_rec3 <- rename(sp_rec3, Snapshot.Name = question__species)

#join by snapshot name  (filter mammals at the end?)

sp_rec4 <- left_join(sp_rec3, traits2, by = "Snapshot.Name") #working


# put together a final df with final cols --------------------------------


# --- Create an empty dataframe with the columns we need
# The columns needed
cols <- c("Institution ID",	"Collection ID",	"Dataset ID",	"Basis of Record",	"Catalogue Number",
          "Occurrence Remarks",	"Record Number",	"Recorded By",	"Individual ID",	"Individual Count",
          "Life Stage",	"Reproductive Condition",	"Establishment Means",	"Preparations",	"Disposition",
          "Associated Reference",	"Associated Sequences",	"Sampling Protocol",	"Sampling Effort",	"Year",
          "Month",	"Day",	"Habitat",	"Field Number",	"Field Notes","Country",	"Province",	"Municipality", "Locality",	"Elevation",	"Location According To",	"Location Remarks",	"Decimal Latitude",	"Decimal Longitude","geodetic Datum",	"Co-ordinate Uncertainty in Meters","IdentifiedBy",	"Date Identified",
          "Identification references",	"Identification Remarks","Identification Qualifier","Type Status",
          "Name according to Id",	"full Scientific Name+author details",	"Kingdom",	"Phylum",	"Class","Order", "Family",	"Genus",	"Species",	"Infraspecific Epithet",	"Scientific Name Authorship",
          "Vernacular Name",	"Taxon Rank",	"Nomenclatural Status",	"Taxon Remarks", "Camera.Site","metadata_Behaviour","metadata_Sex", "metadata_young_present","EXIF.Make","subject_id", "DateTimeOriginal")    
      
ncol <- length(cols) # number of columns

# --- Create columns that don't exist 
# Helper function to add columns that are not in the original data

fncols <- function(data, cname) {
  # data = the dataframe to modify
  # cname = column names (character vector or character)
  res <- data
  add <- cname[!cname %in% names(data)]
  
  if(length(add)!=0){
    res[add] <- NA
  } 
  return(res)
}

#run the function
df_res <- fncols(sp_rec4, cols) 

# --- Reorder and delete cols not needed
df_fin <- df_res %>% select(all_of(cols)) # reorder in the same order as cols used before

#delete NA for vernacular name (those must be birds and others not mammals)

df_fin1 <- df_fin %>% 
  filter(!is.na(`Vernacular Name`))

locality <- df_fin$Locality[1]
myfile <- file.path("data_out", paste0(locality, "_FBIP_Form.csv"))
write.csv(df_fin1, file = myfile) 
return(df_fin1)
}

```

##  so let´s run above function

should create a file with the name of the location

```{r}
repDwC <- DarwinCore(sp_rec,metadata,traits) #should create a file with the name of the location, error is not an issue consider a warning...
#6008-5972 = 36, minus 36 (bycatch) = 0!! now finally this shit isworking, NO missing recs
```


#END-----
